---
title: "Introduction to nonbgfa"
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to nonbgfa}
  %\VignetteEngine{quarto::quarto}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**nonbgfa** identifies sequences associated with non-B DNA-forming motifs in
genomic FASTA sequences. It wraps the GFA suite originally developed at
NCI-Frederick / Frederick National Laboratory for Biodefense and
Counterterrorism.

Seven motif classes are supported:

| Function | Motif | Structure |
|---|---|---|
| `find_ir()` | Inverted repeats | Cruciform DNA |
| `find_mr()` | Mirror repeats | Triplex (H-DNA) |
| `find_dr()` | Direct repeats | Slipped-strand DNA |
| `find_gq()` | G-quadruplexes | Tetraplex (G4 DNA) |
| `find_zdna()` | Z-DNA | Left-handed helix |
| `find_str()` | Short tandem repeats | Microsatellites |
| `find_apr()` | A-phased repeats | Bent DNA |

## Installation

```r
# Install from GitHub (package not yet on CRAN)
# install.packages("pak")
pak::pak("NCATS-Tangerine/non-b-gfa")
```

## Quick start

```{r quick-start, eval=FALSE}
library(nonbgfa)

fasta <- system.file("extdata", "gfa_test.fasta", package = "nonbgfa")

# Run all motif finders at once
results <- find_nonb(fasta)

# Individual motif types
ir  <- find_ir(fasta)
gq  <- find_gq(fasta)
```

## Interpreting results

Each finder returns a `data.frame` with the following columns:

- **seq_name**: Sequence identifier from the FASTA header.
- **start / end**: 1-based genomic coordinates.
- **strand**: `"+"` or `"-"`.
- **length**: Length of the repeat unit or G-run.
- **spacer**: Spacer between repeat halves (or KV score for Z-DNA).
- **num_repeats**: Number of times the unit is repeated.
- **remainder**: Motif-specific (DR remainder, IR/MR min loop, GQ island count).
- **subset**: `TRUE` if the locus qualifies as a special structural subset
  (cruciform, triplex, slipped, or KV-scored Z-DNA).

## Using GenomicRanges output

```{r granges, eval=FALSE}
# Requires Bioconductor: BiocManager::install("GenomicRanges")
ir_gr <- find_ir(fasta, format = "GRanges")
ir_gr
```

## Adjusting search parameters

All parameters mirror the original `gfa` command-line switches and default
to the same values used by the Non-B DB website.

```{r params, eval=FALSE}
# Require longer G-runs and tighter spacers for G-quadruplexes
strict_gq <- find_gq(fasta, minGQrep = 4, maxGQspacer = 3)

# Search only for inverted repeats with longer arms
long_ir <- find_ir(fasta, minIRrep = 12, maxIRspacer = 50)
```

## Citation

Please cite the original GFA tool:

> Cer RZ, Donohue DE, Mudunuri US, Temiz NA, Loss MA, Starner NJ,
> Halusa GN, Volfovsky N, Yi M, Luke BT, Bacolla A, Collins JR,
> Stephens RM. (2013) Non-B DB v2.0: a database of predicted non-B
> DNA-forming motifs and its associated tools. *Nucleic Acids Research*,
> 41(D1):D94–D100. <https://doi.org/10.1093/nar/gks955>
